---
title: "CPSM Workflow: Cancer Patient Survival Modeling"
author: "Harpreet Kaur, Pijush Das, Kevin Camphausen, Uma Shankavaram"
date: "2025-12-26"
output:
  BiocStyle::html_document: default
vignette: >
  %\VignetteIndexEntry{CPSM Workflow: Cancer Patient Survival Modeling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

# Overview

This workflow demonstrates the main CPSM analysis pipeline for cancer patient survival modeling:

1. load example TCGA data distributed with the package,
2. preprocess and split the cohort,
3. normalize gene-expression values,
4. compute LASSO-based prognostic index (PI) scores,
5. run univariate survival feature selection,
6. train MTLR-based survival prediction models,
7. generate patient-level survival plots and risk-group predictions,
8. create a nomogram from selected clinical variables.

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install("CPSM")
```

# Load packages and example data

```{r load-data}
library(CPSM)
library(SummarizedExperiment)
library(dplyr)

set.seed(7)

data(Example_TCGA_LGG_FPKM_data, package = "CPSM")

# samples in rows; clinical columns first, followed by expression columns
combined_df <- cbind(
  as.data.frame(colData(Example_TCGA_LGG_FPKM_data))[ , -ncol(colData(Example_TCGA_LGG_FPKM_data)), drop = FALSE],
  t(as.data.frame(assay(Example_TCGA_LGG_FPKM_data, "expression")))
)

str(combined_df[1:3, 1:10])
```

# Step 1: Data processing

```{r data-processing}
New_data <- data_process_f(combined_df, col_num = 20, surv_time = "OS.time")
str(New_data[1:3, 1:10])
```

# Step 2: Train/test split

```{r train-test-split}
result_split <- tr_test_f(data = New_data, fraction = 0.9)
train_FPKM <- result_split$train_data
test_FPKM  <- result_split$test_data

str(train_FPKM[1:3, 1:10])
str(test_FPKM[1:3, 1:10])
```

# Step 3: Normalization

```{r normalization}
Result_N_data <- train_test_normalization_f(
  train_data = train_FPKM,
  test_data = test_FPKM,
  col_num = 21
)

Train_Clin <- Result_N_data$Train_Clin
Test_Clin <- Result_N_data$Test_Clin
Train_Norm_data <- Result_N_data$Train_Norm_data
Test_Norm_data <- Result_N_data$Test_Norm_data

str(Train_Clin[1:3, 1:10])
str(Test_Clin[1:3, 1:10])
```

# Step 4a: LASSO PI score calculation

```{r lasso-pi}
Result_PI <- Lasso_PI_scores_f(
  train_data = Train_Norm_data,
  test_data = Test_Norm_data,
  nfolds = 5,
  n_repeats = 50,
  alpha = 1,
  col_num = 21,
  freq_threshold = 0.6,
  surv_time = "OS_month",
  surv_event = "OS"
)

Stability_table <- Result_PI$Stability_table
Train_Lasso_key_variables <- Result_PI$Train_Lasso_key_variables
Train_PI_data <- Result_PI$Train_PI_data
Test_PI_data <- Result_PI$Test_PI_data

head(Stability_table[order(-Stability_table$Selection_Frequency), ])
head(Train_Lasso_key_variables)
str(Train_PI_data)
str(Test_PI_data)
```

# Step 4b: Univariate survival feature selection

```{r univariate-selection}
Result_Uni <- Univariate_sig_features_f(
  train_data = Train_Norm_data,
  test_data = Test_Norm_data,
  col_num = 21,
  surv_time = "OS_month",
  surv_event = "OS"
)

Univariate_Suv_Sig_G_L <- Result_Uni$Univariate_Survival_Significant_genes_List
Train_Uni_sig_data <- Result_Uni$Train_Uni_sig_data
Test_Uni_sig_data <- Result_Uni$Test_Uni_sig_data
Uni_Sur_Sig_clin_List <- Result_Uni$Univariate_Survival_Significant_clin_List
ZPH_test <- Result_Uni$ZPH_Genes
PH_Summary <- Result_Uni$PH_Summary_Genes

head(Univariate_Suv_Sig_G_L)
head(Uni_Sur_Sig_clin_List)
head(ZPH_test)
head(PH_Summary)
```

# Step 5: Survival prediction models

## Model 1: Clinical features only

```{r model-clinical}
Result_Model_Type1 <- MTLR_pred_model_f(
  train_clin_data = Train_Clin,
  test_clin_data = Test_Clin,
  Model_type = 1,
  train_features_data = Train_Clin,
  test_features_data = Test_Clin,
  Clin_Feature_List = Key_Clin_feature_list,
  surv_time = "OS_month",
  surv_event = "OS",
  nfolds = 5
)

survCurves_data_clin <- Result_Model_Type1$survCurves_data
mean_median_survival_time_clin <- Result_Model_Type1$mean_median_survival_time_data
Error_mat_for_Model_clin <- Result_Model_Type1$Error_mat_for_Model
head(Error_mat_for_Model_clin["Test_set", ])
```

## Model 2: PI score only

```{r model-pi}
Result_Model_Type2 <- MTLR_pred_model_f(
  train_clin_data = Train_Clin,
  test_clin_data = Test_Clin,
  Model_type = 2,
  train_features_data = Train_PI_data,
  test_features_data = Test_PI_data,
  Clin_Feature_List = Key_PI_list,
  surv_time = "OS_month",
  surv_event = "OS",
  nfolds = 5
)

survCurves_data <- Result_Model_Type2$survCurves_data
mean_median_survival_time_data <- Result_Model_Type2$mean_median_survival_time_data
survival_result_based_on_MTLR <- Result_Model_Type2$survival_result_based_on_MTLR
Fold_Metrics <- Result_Model_Type2$Fold_Metrics
fold_summary_df <- Result_Model_Type2$fold_summary_df
Error_mat_for_Model <- Result_Model_Type2$Error_mat_for_Model

str(survCurves_data)
str(mean_median_survival_time_data)
head(Error_mat_for_Model["Test_set", ])
```

## Model 3: Clinical features + PI

```{r model-clinical-plus-pi}
Result_Model_Type3 <- MTLR_pred_model_f(
  train_clin_data = Train_Clin,
  test_clin_data = Test_Clin,
  Model_type = 3,
  train_features_data = Train_PI_data,
  test_features_data = Test_PI_data,
  Clin_Feature_List = Key_Clin_features_with_PI_list,
  surv_time = "OS_month",
  surv_event = "OS",
  nfolds = 5
)

survCurves_data_3 <- Result_Model_Type3$survCurves_data
mean_median_surv_tim_da_3 <- Result_Model_Type3$mean_median_survival_time_data
Error_mat_for_Model_3 <- Result_Model_Type3$Error_mat_for_Model
head(Error_mat_for_Model_3["Test_set", ])
```

## Model 4: Univariate features + clinical features

```{r model-univariate-plus-clinical}
Result_Model_Type4 <- MTLR_pred_model_f(
  train_clin_data = Train_Clin,
  test_clin_data = Test_Clin,
  Model_type = 4,
  train_features_data = Train_Uni_sig_data,
  test_features_data = Test_Uni_sig_data,
  Clin_Feature_List = Key_univariate_features_with_Clin_list,
  surv_time = "OS_month",
  surv_event = "OS",
  nfolds = 5
)

survCurves_data_4 <- Result_Model_Type4$survCurves_data
Error_mat_for_Model_4 <- Result_Model_Type4$Error_mat_for_Model
head(Error_mat_for_Model_4["Test_set", ])
```

# Step 6: Survival curves for individual patients

```{r survival-curves, fig.width = 7, fig.height = 4}
plots <- surv_curve_plots_f(
  Surv_curve_data = survCurves_data,
  selected_sample = rownames(survCurves_data)[1],
  font_size = 12,
  line_size = 0.5,
  all_line_col = "grey70",
  highlight_col = "red"
)

plots$all_patients_plot
plots$highlighted_patient_plot
```

# Step 7: Predicted mean and median survival time

```{r survival-barplot, fig.width = 7, fig.height = 4}
plots_2 <- mean_median_surv_barplot_f(
  surv_mean_med_data = mean_median_survival_time_data,
  selected_sample = rownames(mean_median_survival_time_data)[1],
  font_size = 8,
  font_color = "black"
)

plots_2$mean_med_all_pat
plots_2$highlighted_selected_pat
```

# Step 8: Risk-group prediction

```{r risk-group}
Results_Risk_group_Prediction <- predict_survival_risk_group_f(
  selected_train_data = Train_PI_data,
  selected_test_data = Test_PI_data,
  Feature_List = Key_PI_list
)

Best_model_Prediction_results <- Results_Risk_group_Prediction$misclassification_results
Test_results <- Results_Risk_group_Prediction$Test_results
Fold_results <- Results_Risk_group_Prediction$CV_Fold_Results

head(Best_model_Prediction_results)
head(Test_results)
```

# Step 9: Kaplan-Meier overlay plot

```{r km-overlay, fig.width = 8, fig.height = 6}
KM_plot <- km_overlay_plot_f(
  Train_results = Train_results,
  Test_results = Test_results,
  survcurve_te_data = survCurves_data,
  selected_sample = rownames(Test_results)[1],
  font_size = 12,
  train_palette = c("firebrick", "blue"),
  test_curve_col = "darkgreen",
  test_curve_size = 1.2,
  test_curve_lty = "dotdash",
  annotation_col = "black"
)

KM_plot
```

# Step 10: Nomogram

```{r nomogram, fig.width = 7, fig.height = 6}
Result_Nomogram <- Nomogram_generate_f(
  data = Train_Data_Nomogram_input,
  Feature_List = feature_list_for_Nomogram,
  surv_time = "OS_month",
  surv_event = "OS",
  font_size = 0.8,
  axis_cex = 0.5,
  tcl_len = 0.5,
  label_margin = 0.5,
  col_grid = gray(c(0.85, 0.95))
)

C_index_mat <- Result_Nomogram$C_index_mat
C_index_mat
```

# Session info

```{r session-info}
sessionInfo()
```
